Sherlock topic model paper

### 2018-05-21 main.tex
saving progress, writing up classic memory analyses

- using list-learning / trial based experiments, we've learned a great deal about the human memory system.
- however, its undeniable that many lab-based experiments are severely contrived as compared to real life experiences
- for example, real life experiences are typically highly autocorrelated in space and time, a feature that is not typically present in traditional laboratory studies.
- nonetheless, we want to leverage the tools and knowledge gained over the past few decades
- here, we present a methodological and theoretical advance in our understanding of memory for naturalistic experiences.
- our approach is model the semantic information contained in a stimulus and project memory responses into this stimulus space, allowing for fine-grained analysis of the mnemonic reprsentations.

@@

Next, we tested the hypothesis that the individual `blocks' along the diagonals of the recall matrices represent the recall of a particular video event. To this end, first we created a video event model, which was done by averaging together neighboring topic vectors that the were classified to be in the same event (\ref{fig:eventseg}c), resulting in an events (34) by topics (100) matrix.  We performed the same procedure for the recall matrices (see \ref{fig:eventseg}d for example). Then, we computed the correlation between these two event models, resulting in a video events (34) by recall events (8-27, depending on the participant) correlation matrix (see \ref{fig:eventseg}e). These matrices represent the extent to which each recall event is similar to each video event (for each participant). To determine which video events the recall events most likely referred to, for each recall event, we found the index of the video event with the highest correlation (i.e. the argmax).  This is depicted in \ref{fig:eventseg}e as the cell highlighted with the black box for each column in the video-recall event correlation matrix. Notably, our algorithm suggests that the example participant recalled the events largely in order, occasionally skipping over an event.

@@

To better visualize the relationship between the video and recall models, we embedded the video and average recall model into a 2D space (using the UMAP dimensionality reduction algorithm, for details see \ref{sec:methods}) where the points represent video/recall events and the distance between them represents their similarity in 'topic space' (\ref{fig:eventseg}g). We computed the group average transition probabilites from each event to each other event and plotted the values as lines where the tranparency is proportional to the probability of the transition.  By eye, it is clear that the two models have a very similar geometric structure, and the transition probabilities tend to be strongest between events that occured sequentially (or nearby) in time.

To explore individual variability (and consistency) in the recall event models, we embedded each participant's recall event model in the same 2D 'topics space' described above (\ref{fig:eventseg}h). Despite the variability in the number of events remembered, the overall shape of most of the participants' recall trajectories are similar to the video model (\ref{fig:eventseg}h), particularly the participants with a greater number of recalled events. To quantify the similarity between the video model and individual recall models, we considered a number of metrics.  First, we tested whether each participant's recall model matched the movie model in a general sense. To do this, for each participant we filtered the video model to only include the events that the participant remembered. Then, we computed the root mean squared difference (RMSD) between the video model and the recall model. As an example, if the participant remembered all the events in order (with perfect precision), the expected distance value would be 0. However, if they remembered a subset of events, events our of order or with low precision the expected distance would be greater than 0. To assess significance, we permformed a permutation test where we circularly shifted the recall model (10000 times) and recomputed the RMSD. The recall model significantly matched the video model for nine of the participants (p<.05; participants: 3-4, 8-13, 17) and the p-value for the rest of the participants was less than .25. Furthermore, the RMSD values were significantly correlated to memory performance across participants (Pearson's r=-.57, p=.016)  Next, we tested whether participants who recalled more events were also more precise in their recollections. For each participant, we computed the correlation between each recall event and its matching video event (only for the events which they recalled) and averaged the correlation within participant.  In line with our prediction, there was a strong correlation between memory performance and precision suggesting that participants who remembered more events also remembered them more veridically (Pearson's r=.74, p=.0006). Related to this last measure, we considered the distinctiveness of each recall event. That is, how unique each recall event is with respect to all other recall events. We hypothesized that participants with high memory performance might describe each event in a more distinctive way (compared to those with lower memory performance who might describe events in a more general way). To this end, we computed a 'distinctiveness' score for each participant, i.e. 1 - correlation between the given recall event and all other recall events for the participant.  Then, we averaged this measure across recall events.  We found that participants with higher memory performance also had higher distinctiveness scores (Pearson's r=.8, p=.0001). Finally, we tested whether participants with better memory performance were more likely to remember the events in order.  For each participant, we computed the Spearman rank correlation between the order of events that the participant recalled and the actual order of events, sampling out only events that were actually recalled.  We found that participants who recalled more events also recalled more of them in order (Pearson's r=.5, p=.04).

\subsubsection{Naturalistic extensions of classic memory analyses}
s
Traditionally, a major limitation of 'naturalistic' memory experiments has been quantifying memory dynamics in a principled way.
In the next section, we extend a set of classic memory analyses that were originally designed to measure memory in 'free recall' list-learning experiments to naturalistic settings. Just like recall

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/3_behav_eventseg.pdf}
\caption{\label{fig:behav}Naturalistic extensions of classic memory analyses. A-C use the `smooth' approach and D-F use the `best' approach (see Methods for details). A/D. The probability of first recall as a function of the serial position the the event during encoding. B/E. Given recall of event i, the probability that the next recalled item will be from serial position i \rpm~lag. C/F. Proportion recalled as a function of serial position. All error bars are the standard error of the mean derived from a bootstrap resampling procedure.
}
\end{figure}

>>>>

- using list-learning / trial based experiments, we've learned a great deal about the human memory system.

- however, its undeniable that many lab-based experiments are severely contrived as compared to real life experiences

- for example, real life experiences are typically highly autocorrelated in space and time, a feature that is not typically present in traditional laboratory studies.

- nonetheless, we want to leverage the tools and knowledge gained over the past few decades

- here, we present a methodological and theoretical advance in our understanding of memory for naturalistic experiences.

- our approach is model the semantic information contained in a stimulus and project memory responses into this stimulus space, allowing for fine-grained analysis of the mnemonic reprsentations.

@@

Next, we tested the hypothesis that the individual `blocks' along the diagonals of the recall matrices represent the recall of a particular video event. To test this, first we created a video event model, which was done by averaging together neighboring topic vectors that the were classified to be in the same event (\ref{fig:eventseg}c), resulting in an events (34) by topics (100) matrix.  We performed the same procedure for the recall matrices (see \ref{fig:eventseg}d for example). Then, we computed the correlation between these two event models, resulting in a video events (34) by recall events (8-27, depending on the participant) correlation matrix (see \ref{fig:eventseg}e). These matrices represent the extent to which each recall event is similar to each video event (for each participant). To determine which video events the recall events most likely referred to, for each recall event, we found the index of the video event with the highest correlation (i.e. the argmax).  This is depicted in \ref{fig:eventseg}e as the cell highlighted with the black box for each column in the video-recall event correlation matrix. Notably, our algorithm suggests that the example participant recalled the events largely in order, occasionally skipping over an event.

@@

To better visualize the relationship between the video and recall models, we embedded the video and average recall model into a 2D space (using the UMAP dimensionality reduction algorithm, for details see \ref{sec:methods}) where the points represent video/recall events and the distance between them represents their similarity in 'topic space' (\ref{fig:eventseg}g). We computed the group average transition probabilites from each event to each other event and plotted the values as lines where the tranparency is proportional to the probability of the transition.  By eye, it is clear that the two models have a very similar geometric structure, and the transition probabilities tend to be strongest between events that occured sequentially (or nearby) in time. To explore individual variability (and consistency) in the recall event models, we embedded each participant's recall event model in the same 2D 'topics space' described above (\ref{fig:eventseg}h). Despite the variability in the number of events remembered, the overall shape of most of the participants' recall trajectories are similar to the video model (\ref{fig:eventseg}h), particularly the participants with a greater number of recalled events.

\subsubsection{Measuring the quality of recall events}
Representing the video and verbal recall as events in `topic space' affords us the ability to characterize the quality of recall in a more fine-grained and nuanced way than what is typically possible. To quantify the similarity between the video model and individual recall models, we considered a number of novel metrics.  First, we tested whether each participant's recall model matched the movie model in a general sense. To do this, for each participant we filtered the video model to only include the events that the participant remembered. Then, we computed the root mean squared difference (RMSD) between the video model and the recall model. As an example, if the participant remembered all the events in order (with perfect precision), the expected distance value would be 0. However, if they remembered a subset of events, events our of order or with low precision the expected distance would be greater than 0. To assess significance, we performed a permutation test where we circularly shifted the recall model (10000 times) and recomputed the RMSD. The recall model significantly matched the video model for nine of the participants (p<.05; participants: 3-4, 8-13, 17 and the p-value for the rest of the participants was less than .25). Furthermore, the RMSD values were significantly correlated to memory performance across participants (Pearson's r=-.57, p=.016). Thus, a closer match between the video and recall event models was related to better recall performance.

Next, we tested whether participants who recalled more events were also more precise in their recollections. For each participant, we computed the correlation between each recall event and its matching video event (only for the events which they recalled). This resulted in a single number for each recalled event indexing how similar the recall event was to its matching movie event (i.e the `precision' of the recall). We then averaged the correlations within participant. In line with our prediction, there was a strong correlation between memory performance and precision suggesting that participants who remembered more events also remembered them more veridically (Pearson's r=.74, p=.0006). Next, we considered the distinctiveness of each recall event. That is, how uniquely a recall event matched a given video event compared to all other video events. We hypothesized that participants with high memory performance might describe each event in a more distinctive way (relative to those with lower memory performance who might describe events in a more general way). To this end, we computed a `distinctiveness' score for each participant (i.e. 1 - the correlation between a recall event and all non-matching video events).  Then, we averaged this measure over recall events within participant.  We found that participants with higher memory performance also had higher distinctiveness scores (Pearson's r=.8, p=.0001). Lastly, we tested whether participants with better memory performance were also more likely to remember the events in order.  For each participant, we computed the Spearman rank correlation between the order of events that the participant recalled and the actual order of events (filtering events that were actually recalled).  We found that participants who recalled more events also recalled more of them in order (Pearson's r=.5, p=.04). In summary, we found that better memory performance was associated with more precise and distinctive recall.

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/3_behav_quality.pdf}
\caption{\label{fig:behav}
}
\end{figure}

\subsubsection{Naturalistic extensions of classic memory analyses}
In addition to the novel metrics we introduce above, our approach also allows us to extend classic memory analyses designed to characterize memory in list-learning experiments to more naturalistic settings. Just like in a `free-recall' list-learning experiment where participants verbally recall a series of unique items in some order, our video-recall matching approach allows to analyze memory in the same way.

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/4_behav_eventseg.pdf}
\caption{\label{fig:behav}Naturalistic extensions of classic memory analyses. A-C use the `smooth' approach and D-F use the `best' approach (see Methods for details). A/D. The probability of first recall as a function of the serial position the the event during encoding. B/E. Given recall of event i, the probability that the next recalled item will be from serial position i \rpm~lag. C/F. Proportion recalled as a function of serial position. All error bars are the standard error of the mean derived from a bootstrap resampling procedure.
}
\end{figure}
